Properties of winning Iterated Prisoner's Dilemma strategies.

### 2020-04-01 paper/main.tex
run regression on median score
minor suggestions by VK
Co-Authored-By: Vince Knight <vincent.knight@gmail.com>

Other works have built upon the limitations of the TFT, and others have shown
that suggestions made by Axelrod did not necessarily lift to alternative environmental rules.
In~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984} it was shown
that TFT suffered in environments with noise. This was
mainly due to the strategy's lack of generosity and contrition. Since TFT immediately
punishes a defection, in a noisy environment it can get stuck in a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced, including
Nice and Forgiving~\cite{Bendor1991}, Generous Tit For Tat~\cite{Nowak1992},
and Pavlov (aka Win Stay Lose Shift)~\cite{Nowak1993}, as well as later
variants such as OmegaTFT \cite{kendall2007iterated}. Introduction of new strategies
is often accompanied by a claim that the new strategy is the best known despite
only being tested against a small number of opponents, or specific classes of
opponents not necessarily representative of all possible or all published strategies.
The lack of testing against formally defined strategies and tournament winners
is understandable given the effort required to implement
\footnote{Implementing prior strategies faithfully is often extremely difficult or impossible
due to insufficient descriptions and lack of published implementations or code.}
the hundreds of published IPD strategies, yet calls into question any claims of superiority
or robustness of newly introduced
strategies.

@@

In comparison, the performance the top ranked strategies in noisy environments
when \(p_n\in [0, 1]\) is bimodal. The top strategies include strategies which
decide their actions based on the cooperation to defection ratio, such as
ShortMem~\cite{Andre2013}, Grumpy~\cite{axelrodproject} and
e~\cite{axelrodproject}, and the Retaliate strategies which are designed to
defect if the opponent has tricked them more often than x\% of the times that
they have done the same. The bimodality of the \(r\) distributions is explained
by Figure~\ref{fig:effect_of_noise} which demonstrates that the top 6 strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly. At a noisy level of \(0.5\) or greater, mostly cooperative strategies
become mostly defectors and vice versa.

@@

\begin{itemize}
    \item In standard tournaments the dominating strategies were clever
    strategies that had been trained using reinforcement learning techniques.
    \item In noisy environments where the noise probability strictly less than
    0.1 was considered, the successful strategies were strategies specifically
    designed for noisy environments.
    \item In probabilistic ending tournaments most of the highly ranked
    strategies were defecting strategies and trained finite state automata, all
    by the authors of~\cite{Ashlock2006, Ashlock2014}. These strategies ranked
    high due to their performance in tournaments where the probability of the
    game ending after each turn was bigger than 0.1.
    \item In probabilistic tournaments with \(p_e\) less than 0.1 the highly
    ranked strategies were strategies based on the behaviour of others.
    \item From the collection of strategies considered here,  no strategy can be
    consistently successful in noisy environments, except if the value of noise
    is constrained to less than a 0.1.
\end{itemize}

@@

Now we examine performance of the strategies based on features of strategies described in
Table~\ref{table:manual_features}. These feature are measures regarding a
strategy's behaviour from the tournaments the strategies competed in as well as
intrinsic properties such as whether a strategy is deterministic or stochastic.

@@

Similarly, these results suggest an explanation regarding the perhaps intuitively unexpected
effectiveness of memory one strategies historically. Given that among the important
features associated to success are the relative cooperation rate to the population
average and the four memory-one probabilities of cooperating conditional on the
previous round of play, all five features can be optimized by a memory one
strategy such as TFT. Usage of more history becomes valuable when there are
exploitable opponent patterns, indicated by the importance of SSE as a feature,
that the first-approximation provided by a memory one strategy is no longer
sufficient.

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
\resizebox{.8\textwidth}{!}{
    \input{regression_table.tex}}
    \end{center}
    \caption{Results of multivariate linear regressions with \(r\) as the dependent variable.
    \(R\) squared is reported for each model.}
    \label{table:linear_regression}
\end{table}

@@

The distributions of SSE and \(CC\) to \(D\) rate for the winners of standard
tournaments are also given in Figure~\ref{fig:discussion_standard}. The SSE
distributions for the winners indicate that they strategy's behaved in ZD way in
several tournaments, however, not constantly. The winners participated in
matches where they did not try to extortionate their opponents. Furthermore, the
\(CC\) to \(D\) distribution indicates that if a strategy were to defect against
the winners they would reciprocate on average with a probability of 0.5.

@@

A third method that evaluates the importance of the features in
Table~\ref{table:manual_features} using clustering and random forests can be found in
the Appendix~\ref{app:clustering}. The results uphold the outcomes of the
correlation and multivariate regression methods. It also evaluates the effects
of the classifiers stochastic, make use of game, and make use of length which
have not been evaluated by the methods above because there are binary variables.
The results imply that they have no significant effect on a strategy's
performance.

@@

So what is the best way of playing the IPD? And is there a single dominant
strategy for the IPD? There was not a single strategy within the collection of the
\numberofstrategies strategies, that has managed to perform well in all the
tournaments variations it competed in.
Even if on average a strategy ranked highly in a specific environment it did not
guarantee its success over the different tournament types. However, the results
of sections~\ref{section:top_performances}
and~\ref{section:evaluation_of_performance} have demonstrated that there are
properties associated with the success of strategies. A few the properties that
have been identified by this manuscript's analysis contradict the properties of
Axelrod~\cite{Axelrod1981}. Namely, in Section~\ref{section:top_performances} it
was shown that trained strategies and strategies that decided their actions
based on pre-designed strategies to maximise their utility dominated several
tournaments across tournament types, hinting that successful IPD strategies are
often clever or more complex than simple strategies like TFT. Most of the
successful strategies highlighted
in~Section~\ref{section:top_performances} were strategies that begin with
cooperation.

@@

\input{parameters_section}
\input{correlation_section}
\input{random_forest_evaluation}
\input{strategies_list_section}

>>>>

Other works have built upon the limitations of TFT, and others have shown
that suggestions made by Axelrod did not necessarily lift to alternative environmental rules.
In~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984} it was shown
that TFT suffered in environments with noise. This was
mainly due to the strategy's lack of generosity and contrition. Since TFT immediately
punishes a defection, in a noisy environment it can get stuck in a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced, including
Nice and Forgiving~\cite{Bendor1991}, Generous Tit For Tat~\cite{Nowak1992},
and Pavlov (aka Win Stay Lose Shift)~\cite{Nowak1993}, as well as later
variants such as OmegaTFT \cite{kendall2007iterated}. Introduction of new strategies
is often accompanied by a claim that the new strategy is the best known despite
only being tested against a small number of opponents, or specific classes of
opponents not necessarily representative of all possible or all published strategies.
The lack of testing against formally defined strategies and tournament winners
is understandable given the effort required to implement
\footnote{Implementing prior strategies faithfully is often extremely difficult or impossible
due to insufficient descriptions and lack of published implementations or code.}
the hundreds of published IPD strategies, yet calls into question any claims of superiority
or robustness of newly introduced
strategies.

@@

In comparison, the performance of the top ranked strategies in noisy environments
when \(p_n\in [0, 1]\) is bimodal. The top strategies include strategies which
decide their actions based on the cooperation to defection ratio, such as
ShortMem~\cite{Andre2013}, Grumpy~\cite{axelrodproject} and
e~\cite{axelrodproject}, and the Retaliate strategies which are designed to
defect if the opponent has tricked them more often than x\% of the times that
they have done the same. The bimodality of the \(r\) distributions is explained
by Figure~\ref{fig:effect_of_noise} which demonstrates that the top 6 strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly. At a noisy level of \(0.5\) or greater, mostly cooperative strategies
become mostly defectors and vice versa.

@@

\begin{itemize}
    \item In standard tournaments the dominating strategies were 
    strategies that had been trained using reinforcement learning techniques.
    \item In noisy environments where the noise probability strictly less than
    0.1 was considered, the successful strategies were strategies specifically
    designed for noisy environments.
    \item In probabilistic ending tournaments most of the highly ranked
    strategies were defecting strategies and trained finite state automata, all
    by the authors of~\cite{Ashlock2006, Ashlock2014}. These strategies ranked
    high due to their performance in tournaments where the probability of the
    game ending after each turn was bigger than 0.1.
    \item In probabilistic tournaments with \(p_e\) less than 0.1 the highly
    ranked strategies were strategies based on the behaviour of others.
    \item From the collection of strategies considered here,  no strategy can be
    consistently successful in noisy environments, except if the value of noise
    is constrained to less than a 0.1.
\end{itemize}

@@

Now we examine performance of the strategies based on features of strategies described in
Table~\ref{table:manual_features}. These features are measures regarding a
strategy's behaviour from the tournaments the strategies competed in as well as
intrinsic properties such as whether a strategy is deterministic or stochastic.

@@

Similarly, these results suggest an explanation regarding the intuitively unexpected
effectiveness of memory one strategies historically. Given that among the important
features associated with success are the relative cooperation rate to the population
average and the four memory-one probabilities of cooperating conditional on the
previous round of play, all five features can be optimized by a memory one
strategy such as TFT. Usage of more history becomes valuable when there are
exploitable opponent patterns, indicated by the importance of SSE as a feature,
that the first-approximation provided by a memory one strategy is no longer
sufficient.

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
\resizebox{\textwidth}{!}{
    \input{regression_table.tex}}
    \end{center}
    \caption{Results of multivariate linear regressions with \(r\) as the dependent variable.
    \(R\) squared is reported for each model.}
    \label{table:linear_regression}
\end{table}

A multivariate linear regression has also be fitted on the median score. The
coefficients and \(p\) values of the features can be found in
Appendix~\ref{app:regression_median_score}. The results of the two methods
are in agreement.

@@

The distributions of SSE and \(CC\) to \(D\) rate for the winners of standard
tournaments are also given in Figure~\ref{fig:discussion_standard}. The SSE
distributions for the winners indicate that the strategy behaved in a ZD way in
several tournaments, however, not constantly. The winners participated in
matches where they did not try to extortionate their opponents. Furthermore, the
\(CC\) to \(D\) distribution indicates that if a strategy were to defect against
the winners they would reciprocate on average with a probability of 0.5.

@@

A third method that evaluates the importance of the features in
Table~\ref{table:manual_features} using clustering and random forests can be found in
the Appendix~\ref{app:clustering}. The results uphold the outcomes of the
correlation and multivariate regression. It also evaluates the effects
of the classifiers stochastic, make use of game, and make use of length which
have not been evaluated by the methods above because there are binary variables.
The results imply that they have no significant effect on a strategy's
performance.

@@

So what is the best way of playing the IPD? And is there a single dominant
strategy for the IPD? There was not a single strategy within the collection of the
\numberofstrategies strategies, that has managed to perform well in all the
tournaments variations it competed in.
Even if on average a strategy ranked highly in a specific environment it did not
guarantee its success over the different tournament types. However, the results
of sections~\ref{section:top_performances}
and~\ref{section:evaluation_of_performance} have demonstrated that there are
properties associated with the success of strategies. A few of the properties that
have been identified by this manuscript's analysis contradict the properties of
Axelrod~\cite{Axelrod1981}. Namely, in Section~\ref{section:top_performances} it
was shown that trained strategies and strategies that decided their actions
based on pre-designed strategies to maximise their utility dominated several
tournaments across tournament types, hinting that successful IPD strategies are
often clever or more complex than simple strategies like TFT. Most of the
successful strategies highlighted
in~Section~\ref{section:top_performances} were strategies that begin with
cooperation.

@@

\input{parameters_section}
\input{correlation_section}
\input{random_forest_evaluation}
\input{regression_on_median_score}
\input{strategies_list_section}
